Plant occurrence subset by species with complete trait records

### 2024-06-12 code/L1/L1_P6_TropicalAndes_plant_occ.Rmd
add summary code and visualize taxonomic coverage

---
title: "Plant occurrence subset by species with complete trait records"
author: "Hazel J. Anderson"
project: "Plant-Frugivore Diversity"
collaborators: "Beth E. Gerstner, Phoebe L. Zarnetske"
overview: "This script subsets plant occurrence data to species with complete trait coverage."
data input: "ADD"
data output: "ADD"
date: "2023-10-04"
output: html_document
---

# Set file paths
```{r}
data_path_L1 <-file.path('G:/Shared drives/SpaCE_Lab_FRUGIVORIA/data/plants/L1')
output_path_L1 <- file.path('G:/Shared drives/SpaCE_Lab_FRUGIVORIA/data/plants/L1')
```

# Load required packages
```{r}
library(dplyr)
#run if not installed remotes::install_github("FRBCesab/funbiogeo")
library(funbiogeo)
```

# Read in data
```{r}
plant_traits <- read.csv(file.path(data_path_L1, file = "TropicalAndes_imputed_plant_traits.csv"))
plant_occ <- read.csv(file.path(data_path_L1, file = "TropicalAndes_GBIF_plant_occ_cleaned.csv"))
```

# Check trait coverage
```{r}
summary(plant_traits)
```

```{r}
fb_plot_number_species_by_trait(plant_traits)
```

```{r}
Plant_complete <- fb_filter_species_by_trait_coverage(plant_traits, threshold_traits_proportion =  1)
```

# Create list of species with complete trait records
```{r}
plant_species <- unique(Plant_complete$species)
```

# Subset frugivore occurrence data by species list
```{r}
plant_occ_subset <- plant_occ %>%
  filter(species %in% plant_species)
```

@@

```{r}
source("~/GitHub/neotropical_plants/functions/data_summary.R")
```

```{r}
data_summary(plant_occ_subset, plant_occ_subset$species, plant_occ_subset$genus, plant_occ_subset$family)
```

# Write data to csv
```{r}
write.csv(plant_occ_subset, file.path(output_path_L1,"TropicalAndes_GBIF_plant_occ_cleaned_subset.csv"))
```

>>>>

---
title: "Plant occurrence subset by species with complete trait records"
author: "Hazel J. Anderson"
project: "Plant-Frugivore Diversity"
collaborators: "Beth E. Gerstner, Phoebe L. Zarnetske"
overview: "This script subsets plant occurrence data to species with complete trait coverage and summarizes the taxonomic coverage."
data input: "TropicalAndes_imputed_plant_traits.csv, TropicalAndes_GBIF_plant_occ_harmonized_subset.csv, TropicalAndes_plant_checklist.csv"
data output: "TropicalAndes_GBIF_plant_occ_harmonized_subset_final.csv"
date: "2023-10-04"
output: html_document
---

# Set file paths
```{r}
data_path_L0 <-file.path('G:/Shared drives/SpaCE_Lab_FRUGIVORIA/data/plants/L0')
data_path_L1 <-file.path('G:/Shared drives/SpaCE_Lab_FRUGIVORIA/data/plants/L1')
output_path_L1 <- file.path('G:/Shared drives/SpaCE_Lab_FRUGIVORIA/data/plants/L1')
figure_path <- file.path('G:/Shared drives/SpaCE_Lab_FRUGIVORIA/data/plants/figures')

```

# Load required packages
```{r}
library(GIFT)
library(dplyr)
library(sf)
library(tidyr)
library(ggplot2)
```

# Read in data
```{r}
plant_traits <- read.csv(file.path(data_path_L1, file = "TropicalAndes_imputed_plant_traits.csv"))
plant_occ <- read.csv(file.path(data_path_L1, file = "TropicalAndes_GBIF_plant_occ_harmonized_subset.csv"))

# Load Tropical Andes Forest shape
TropicalAndes_IUCNHabitat_Forest <- read_sf(file.path(data_path_L0, "Forest_sf.shp"), layer = "Forest_sf")
```

# Create list of species with complete trait records
```{r}
plant_species <- unique(plant_traits$species)
length(plant_species)
```

# Subset frugivore occurrence data by species list
```{r}
plant_occ_subset <- plant_occ %>%
  filter(species %in% plant_species)
nrow(plant_occ_subset)
```

# Get a checklist of Angiosperms (flowering plants) in the Tropical Andes Forest shapefile
```{r}
checklist <- GIFT_checklists(taxon_name = "Angiospermae", taxonomic_group = TRUE, floristic_group = "all", shp = TropicalAndes_IUCNHabitat_Forest, overlap = "extent_intersect")
```
## Add genus based on work_ID
```{r}
full <- GIFT_taxonomy()
```

```{r}
genus_ID <- checklist$checklists$genus_ID
genus_list <- full %>% filter(taxon_ID %in% genus_ID) %>% select(taxon_ID, taxon_name)
names(genus_list) <- c("genus_ID", "genus")
```

```{r}
checklist_tax <- dplyr::inner_join(checklist$checklists, genus_list)
```

```{r}
reference_ID_list <- unique(checklist_tax$ref_ID)
ref <- GIFT_references()
checklist_ref <- ref[which(ref$ref_ID %in% reference_ID_list),
           c("ref_ID", "ref_long", "geo_entity_ref")]
checklist_ref %>%
  mutate(ref_long = iconv(ref_long, from = "UTF-8", to = "ASCII//TRANSLIT"))
```

@@

```{r}
source("~/GitHub/neotropical_plants/code/data_summary.R")
```

```{r}
data_summary(plant_occ_subset, plant_occ_subset$Accepted_species, plant_occ_subset$genus, plant_occ_subset$family)
```

```{r}
data_summary(checklist_tax, checklist_tax$work_species, checklist_tax$genus, checklist_tax$family)
```

# Visualize taxonomic coverage
```{r}
GBIF_occ_summary <- plant_occ_subset %>%
  summarise(
    num_species = n_distinct(Accepted_species),
    num_families = n_distinct(family),
    num_genera = n_distinct(genus)
  )
GIFT_checklist_summary <- checklist_tax %>%
  summarise(
    num_species = n_distinct(work_species),
    num_families = n_distinct(family),
    num_genera = n_distinct(genus)
  )
```

```{r}
# Combine the summaries into one dataframe
combined_summary <- bind_rows(
  GBIF_occ_summary %>% mutate(source = "Data"),
  GIFT_checklist_summary %>% mutate(source = "Checklist")
)

# Reshape to long format for easier plotting
combined_summary_long <- combined_summary %>%
  pivot_longer(cols = c(num_species, num_families, num_genera), 
               names_to = "category", 
               values_to = "count")
```

```{r}
ggplot(combined_summary_long, aes(x = source, y = count, fill = source)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Comparison of Species, Genera, and Families",
       x = "Source",
       y = "Count",
       fill = "Source") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),  # Angled x-axis labels
    plot.title = element_text(hjust = 0.5),             # Center the plot title
    plot.margin = margin(t = 10, r = 10, b = 10, l = 10)
  ) +
  facet_wrap(~ category, scales = "free_y") +
  theme(strip.text.x = element_text(size = 12, face = "bold"))
ggsave("plant_taxonomic_coverage.png", plot = last_plot(), path = figure_path)
```

# Write data to csv
```{r}
write.csv(plant_occ_subset, file.path(output_path_L1,"TropicalAndes_GBIF_plant_occ_harmonized_subset_final.csv"))
```
